A Way Forward in Parallelising Dynamic Languages

### 2014-05-05 talk/icooolps2014/position-paper.tex
a few tweaks, a few places where more citations are needed

\usepackage[utf8]{inputenc}
\usepackage{array}
\usepackage{hyperref}
\usepackage{amsmath}

@@

Since the GIL is mostly an implementation detail of the interpreter,
it is not exposed to the application running on top of it. To
synchronise memory accesses in applications using threads, the
state-of-the-art still means explicit locking everywhere. It is well
known that using locks for synchronisation is not easy.  They are
non-composable, have overhead, may deadlock, limit scalability, and
overall add a lot of complexity. For a better parallel programming
model for dynamic languages, we propose another, well-known
synchronisation mechanism called \emph{atomic blocks}.

@@

\begin{description}
\item[Performance:] How well does the approach perform compared to the
  GIL on single and multiple threads?
\item[Existing applications:] How big are the changes required to
  integrate with and parallelise existing applications?
\item[Better synchronisation:] Does the approach enable better
  synchronisation mechanisms for applications (e.g. atomic blocks)?
\item[Implementation:] How difficult is it to implement the approach
  in the interpreter?
\item[External libraries:] Does the approach allow for easy
  integration of external libraries?
\end{description}

@@

%% - support of atomic blocks?\\
%% - hard to get right (deadlocks, performance, lock-granularity)\\
%% - very hard to get right for a large language\\
%% - hard to retro-fit, as all existing code assumes GIL semantics\\
%% - (there are some semantic differences, right? not given perfect lock-placement, but well)
%% ( http://www.jython.org/jythonbook/en/1.0/Concurrency.html )

@@

TM can be implemented in software (STM) or in hardware (HTM). There
are also hybrid approaches, which combine the two. We count these
hybrid approaches as STM, since they usually provide the same
capabilities as software-only approaches but with different
performance characteristics. We will now first look at HTM, which
recently gained a lot of popularity by its introduction in common
desktop CPUs from Intel (Haswell generation).

@@

However, STM compared to HTM does not suffer from the same restricting
limitations. Transactions can be arbitrarily long.  This makes it
possible to actually expose transactions to the application in the
form of atomic blocks. This is the only approach that enables a better
synchronisation mechanism than locks for applications \emph{and} still
parallelises when using it. We think this is a very important point
because it not only gives dynamic languages the ability to parallelise
(already commonplace in most other languages), but also pushes
parallel programming forward. Together with sequential consistency it
provides a lot of simplification for parallel applications.

>>>>

\usepackage[utf8]{inputenc}
\usepackage{array}
\usepackage{color}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}

\newcommand{\mynote}[2]{%
  \textcolor{red}{%
    \fbox{\bfseries\sffamily\scriptsize#1}%
    {\small$\blacktriangleright$\textsf{\emph{#2}}$\blacktriangleleft$}%
  }%
}

\newcommand\cfbolz[1]{\mynote{Carl Friedrich}{#1}}

@@

Since the GIL is mostly an implementation detail of the interpreter,
it is not exposed to the application running on top of it. To
synchronise memory accesses in applications using threads, the
state-of-the-art still means explicit locking everywhere. It is well
known that using locks for synchronisation is not easy\cfbolz{citation needed
:-). would be cool if you could find something}.  They are
non-composable, have overhead, may deadlock, limit scalability, and
overall add a lot of complexity\cfbolz{same here, really}. For a better parallel programming
model for dynamic languages, we propose another, well-known
synchronisation mechanism called \emph{atomic blocks}\cfbolz{and here}.

@@

\begin{description}
\item[Performance:] How well does the approach perform compared to the
  GIL on a single and on multiple threads?
\item[Existing applications:] How big are the changes required to
  integrate with and parallelise existing applications?
\item[Better synchronisation:] Does the approach enable better
  synchronisation mechanisms for applications (e.g. atomic blocks)?
\item[Implementation:] How difficult is it to implement the approach
  in the interpreter?
\item[External libraries:] Does the approach allow for easy
  integration of external libraries?
\end{description}

@@

\cfbolz{I think you should mention the commented out point below, that a lot of existing code contains latent races / deadlocks that are just not exposed in a GIL-full world}
%% - support of atomic blocks?\\
%% - hard to get right (deadlocks, performance, lock-granularity)\\
%% - very hard to get right for a large language\\
%% - hard to retro-fit, as all existing code assumes GIL semantics\\
%% - (there are some semantic differences, right? not given perfect lock-placement, but well)
%% ( http://www.jython.org/jythonbook/en/1.0/Concurrency.html )

@@

TM can be implemented in software (STM) or in hardware (HTM). There
are also hybrid approaches, which combine the two. We count these
hybrid approaches as STM, since they usually provide the same
capabilities as software-only approaches but with different
performance characteristics. We will now first look at HTM, which
recently gained a lot of popularity by its introduction in common
desktop CPUs from Intel (Haswell generation).\cfbolz{citation}

@@

However, STM compared to HTM does not suffer from the same restricting
limitations. Transactions can in principle be arbitrarily long.  This makes it
possible to actually expose transactions to the application in the
form of atomic blocks. This is the only approach that enables a better
synchronisation mechanism than locks for applications \emph{and} still
parallelises when using it. We think this is a very important point
because it not only gives dynamic languages the ability to parallelise
(already commonplace in most other languages), but also pushes
parallel programming forward. Together with sequential consistency it
provides a lot of simplification for parallel applications.
